---
title: "DEWSeq Analysis"
author: "Thomas Schwarzl and Sudeep Sahadevan"
date: "07/10/2020"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    toc_depth: 5
params:
  protein: ""
  sampleinfo_file: ""
  countmatrix_file: ""
  annotation_file: ""
  output_windows_file: ""
  output_regions_file: ""
  output_bed_file: ""
  output_Rdata: ""
  min_count: 2
  min_sample: 2
  LRT: FALSE
  p_value_cutoff: 0.05
  lfc_cutoff: 1
  overlap_correction: FALSE
  IHW: TRUE
  decide_fit: TRUE
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r assignData, echo=FALSE, eval=TRUE}
protein <- params$protein
sampleinfo_file <- params$sampleinfo_file
countmatrix_file <- params$countmatrix_file
annotation_file <- params$annotation_file
output_windows_file <- params$output_windows_file
output_regions_file <- params$output_regions_file
output_bed_file <- params$output_bed_file
output_Rdata <- params$output_Rdata
minCount <- params$min_count
minSample <- params$min_sample
p_value_cutoff <- params$p_value_cutoff
lfc_cutoff <- params$lfc_cutoff
```

Sanity check input parameter values

```{r sanityCheck, echo=FALSE, eval=TRUE}
# round of minCount and minSample values first
message("Any decimals given as values for min_count and min_sample parameters will be rounded off to the nearest integer.")
minCount <- round(minCount)
minSample <- round(minSample)
# receive and sanity check p_value_cutoff
if(p_value_cutoff<=0 || p_value_cutoff>=1){
  warning("p_value_cutoff must satisfy: 0<=p_value_cutoff<=1. Resetting to default value: 0.05")
  p_value_cutoff <- 0.05
}
# sanity check log2Foldchange cutoff
if(lfc_cutoff<0){
  warning("lfc_cutoff must be a value >=0. Resetting to default value: 1.00")
  lfc_cutoff <- 1.0
}
# sanity check LRT vs Wald
if(is(params$LRT,'logical')){
  LRT <- params$LRT
}else{
  warning("LRT must be TRUE or FALSE, setting this parameter to default: FALSE")
  LRT <- FALSE
}
# sanity check overlap correction parameter
if(is(params$overlap_correction,'logical')){
  overlap_correction <- params$overlap_correction
}else{
  warning("overlap_correction must be TRUE or FALSE, setting this parameter to default: TRUE")
  overlap_correction <- TRUE
}
# sanity check IHW vs BH correction parameter
if(is(params$IHW,'logical')){
  IHW <- params$IHW
}else{
  warning("IHW must be TRUE or FALSE, setting this parameter to default: TRUE")
  IHW <- TRUE
}
# sanity check automated fit vs parametric fit paramter
if(is(params$decide_fit,'logical')){
  decide_fit <- params$decide_fit
}else{
  warning("decide_fit must be TRUE or FALSE, setting this parameter to default: TRUE")
  decide_fit <- TRUE
}
```

# eCLIP analysis of `r protein` 


## Setup

This is the analysis of __`r protein`__ with    
sampleinfo file: ``r sampleinfo_file``   
countmatrix file: ``r countmatrix_file`` and     
annotation file: ``r annotation_file`` 

with the following threshold:

minimum read count per window per sample: ``r minCount``
number of samples with minimum read count per window: ``r minSample``

using the following parameters:   
p-value cut-off: ``r p_value_cutoff``  
Log2FoldChange cut-off: ``r lfc_cutoff``
use automated method for dispersion estmation: ``r decide_fit``    
use LRT test : ``r LRT``  
use overlap correction: ``r overlap_correction``   
use IHW for FDR correction: ``r IHW``   


```{r check if files exists, echo=FALSE}
checkIfExists <- function(x) { 
  if(!file.exists(x))
    stop(paste0("'", x, "' - file does not exist."))
}

checkIfExists(countmatrix_file)
checkIfExists(annotation_file)
checkIfExists(sampleinfo_file)
```

First, we load the libraries.

```{r load DEWSeq}
requiredPackages <- c('DEWSeq','data.table','IHW','R.utils','tidyverse')
installedPackages <- installed.packages()[,1]
diffPackages <- setdiff(requiredPackages,installedPackages)
if(length(diffPackages)!=0){
  stop('Found missing dependencies! Please install the following package(s): ',paste(diffPackages,collapse = ", "))
}
suppressPackageStartupMessages({
  require(DEWSeq)
  require(tidyverse)
  require(data.table)
  require(IHW)
  require(R.utils)
})
```

## Read in data

Here we read in the window counts

```{r read window counts}
WINDOWCOUNTS <- fread(countmatrix_file, sep = "\t",stringsAsFactors = FALSE) %>% as.data.frame()
rownames(WINDOWCOUNTS) <- WINDOWCOUNTS[,1]
WINDOWCOUNTS <- WINDOWCOUNTS[,-1]
```

and the sample info file

```{r read sample info}
SAMPLEINFO <- read.table(sampleinfo_file,sep="\t",stringsAsFactors = FALSE)
if(ncol(SAMPLEINFO)<2){
  stop("sampleinfo_file ",sampleinfo_file," MUST have atleast two columns: first column should be the sample names used in ",countmatrix_file,
       " and second column must be the experiment type: IP or SMI")
}else if(ncol(SAMPLEINFO)>2){
  message("Found ",ncol(SAMPLEINFO)," columns in ",sampleinfo_file," using the first column as sample name and second column as experiment name")
  SAMPLEINFO <- SAMPLEINFO[,c(1,2)]
}
colnames(SAMPLEINFO) <- c("samples","type")
rownames(SAMPLEINFO) <- SAMPLEINFO[,1]
```

Now we make sure that the sampleinfo file contains the column "type" with values "SMI" and "IP" only. 

```{r sampleSanity}
# make sure that SAMPLEINFO rows and WINDOWCOUNTS columns are in same order
commonSamples <- sort(intersect(colnames(WINDOWCOUNTS),rownames(SAMPLEINFO)))
if(length(commonSamples)!=ncol(WINDOWCOUNTS)){
  stop("The number of samples in ",countmatrix_file," and ",sampleinfo_file," do not MATCH!")
}
SAMPLEINFO <- SAMPLEINFO[ commonSamples, ]
WINDOWCOUNTS <- WINDOWCOUNTS[,commonSamples]
# Now make sure that SAMPLEINFO$type contains only "IP" and "SMI"
typeCheck <- setdiff(unique(SAMPLEINFO$type),c("IP","SMI"))
if(length(typeCheck)!=0){
  stop("The second column in ",sampleinfo_file," should contain analysis types: 'IP' or 'SMI' only. Found unknown value(s): ",
       paste(typeCheck,collapse=", "))
}
```

We make sure that only IP and SMI are in the right factor level order

```{r sampleFactor}
SAMPLEINFO <- SAMPLEINFO %>% mutate(type = factor(type, levels = c("SMI", "IP")))
```



We create the DEWSeq object

```{r dewseqInit}
ddw <- DESeqDataSetFromSlidingWindows(countData  = WINDOWCOUNTS,
                                      colData    = SAMPLEINFO,
                                      annotObj   = annotation_file,
                                      tidy       = FALSE,
                                      design     = ~type)
```


## Prefiltering

```{r prefiltering1}
# remove all empty windows
keep <- rowSums(counts(ddw)) >= 1
ddw <- ddw[keep,]
```


## Estimating size factors


```{r size factors}
ddw <- estimateSizeFactors(ddw)
sizeFactors(ddw)
```

### estimate size factors for only protein_coding genes

```{r protein_coding_size_factors}
ddw_mRNAs <- ddw[ rowData(ddw)[,"gene_type"] == "protein_coding", ]
ddw_mRNAs <- estimateSizeFactors(ddw_mRNAs)
```

### estimate size factors without significant windows

```{r size_factors_no_sig_windows}
ddw_tmp <- ddw
ddw_tmp <- estimateDispersions(ddw_tmp, fitType = "local", quiet = TRUE)
if(LRT){
  ddw_tmp <- nbinomLRT(ddw_tmp,full = ~type,reduced = ~1)
}else{
  ddw_tmp <- nbinomWaldTest(ddw_tmp)
}

tmp_significant_windows <- 
                results(ddw_tmp,
                    contrast = c("type", "IP", "SMI"),
                    tidy = TRUE,
                    filterFun = ihw) %>% 
                dplyr::filter(padj < p_value_cutoff) %>% 
                .[["row"]]
rm(ddw_tmp)
```

estimate the size factors without the significant windows.

```{r final_size_factors}
ddw_mRNAs <- ddw_mRNAs[ !rownames(ddw_mRNAs) %in% tmp_significant_windows, ]
ddw_mRNAs <- estimateSizeFactors(ddw_mRNAs)
```

before thresholding:
```{r threshold1}
dim(ddw)
```

Now threshold the windows read count table.
```{r threshold2}
keep_exp <-  which(rowSums(counts(ddw)>minCount)>=minSample)
ddw <- ddw[keep_exp,]
```

after thresholding:
```{r threshold3}
dim(ddw)
```
assign size factors

```{r final_assign}
sizeFactors(ddw) <- sizeFactors(ddw_mRNAs)
rm( list = c("tmp_significant_windows", "ddw_mRNAs"))
sizeFactors(ddw)
```


## Differential window analysis

### Dispersion estimates

```{r source, echo = FALSE, eval = FALSE}
# source: https://support.bioconductor.org/p/81094/
```
We fit parametric and local fit, and decide the best fit following this [Bioconductor post](https://support.bioconductor.org/p/81094/)

```{r parametric_dispersion}
parametric_ddw  <- estimateDispersions(ddw, fitType="parametric")
if(decide_fit){
  local_ddw  <- estimateDispersions(ddw, fitType="local")
}

```

This is the dispersion estimate for parametric fit
```{r plot parametric fit, fig.wide=TRUE}
plotDispEsts(parametric_ddw, main="Parametric fit")
```

This is the dispersion estimate for local fit, given automated decision fitting is enabled:
```{r plot local fit, fig.wide = TRUE}
if(decide_fit){
  plotDispEsts(local_ddw, main="Local fit")
}

```

This will get the residuals for either fit, only for automated decision fitting
```{r residual fit}
parametricResid <- na.omit(with(mcols(parametric_ddw),abs(log(dispGeneEst)-log(dispFit))))
if(decide_fit){
  localResid <- na.omit(with(mcols(local_ddw),abs(log(dispGeneEst)-log(dispFit))))
  residDf <- data.frame(residuals=c(parametricResid,localResid),fitType=c(rep("parametric",length(parametricResid)),rep("local",length(localResid))))
  summary(residDf)
}

```

and we plot histograms of the fits

```{r plot residual histograms, fig.wide = TRUE}
if(decide_fit){
  ggplot(residDf, aes(x = residuals, fill = fitType)) + scale_fill_manual(values = c("darkred", "darkblue")) + geom_histogram(alpha = 0.5, position='identity', bins = 100) + theme_bw()
}

```

Now, we will decide for the better fit based on median

```{r choose_fit}
summary(parametricResid)
if(decide_fit){
  summary(localResid)
  if (median(localResid) <= median(parametricResid)){
    cat("chosen fitType: local")
    ddw <- local_ddw
  }else{
    cat("chosen fitType: parametric")
    ddw <- parametric_ddw
  }
  rm(local_ddw,parametric_ddw,residDf,parametricResid,localResid)
}else{
  ddw <- parametric_ddw
  rm(parametric_ddw)
}

```



### Wald test or LRT

```{r wald_or_LRT}
if(LRT){
  ddw <- nbinomLRT(ddw,full = ~type, reduced = ~1)
}else{
  ddw <- nbinomWaldTest(ddw)
}

```


### Significance testing

```{r sig_windows}
resultWindows <- resultsDEWSeq(ddw,
                              contrast = c("type", "IP", "SMI"),
                              tidy = TRUE) %>% as_tibble

resultWindows
```


### Multiple hypothesis correction with IHW

You might be interested to correct for multiple hypothesis testing with IHW.

Decide on overlap correction based on the parameter `overlap_correction`

````{r ihw}
if(overlap_correction & IHW){
  resultWindows[,"p_adj_IHW"] <- adj_pvalues(ihw(pSlidingWindows ~ baseMean, 
                     data = resultWindows,
                     alpha = p_value_cutoff,
                     nfolds = 10))
  padjCol <- "p_adj_IHW"
}else if(!overlap_correction & IHW){
  resultWindows[,"p_adj_IHW"] <- adj_pvalues(ihw(pvalue ~ baseMean, 
                     data = resultWindows,
                     alpha = p_value_cutoff,
                     nfolds = 10))
  padjCol <- "p_adj_IHW"
}else if(overlap_correction & !IHW){
  padjCol <- "pSlidingWindows.adj"
}else{
  resultWindows[,'p_adj'] <- p.adjust(resultWindows$pvalue,method="BH")
  padjCol <- 'p_adj'
}

```

Determine significant windows

```{r filter_sig_windows}
resultWindows <- resultWindows %>% 
                      mutate(significant = resultWindows[ ,padjCol] < p_value_cutoff)
sigWindows <- sum(resultWindows$significant)
```


`r sigWindows` windows are significant

```{r sig_windows_show}
resultWindows %>%
   filter(significant) %>% 
   arrange(desc(log2FoldChange)) %>% 
   .[["gene_name"]] %>% 
   unique %>% 
   head(20)
```

### Combining windows

```{r, reg1, message=FALSE, eval=TRUE,include=FALSE}
if(sigWindows>0){
	resultRegions <- extractRegions(windowRes  = resultWindows, padjCol    = padjCol, padjThresh = p_value_cutoff, log2FoldChangeThresh = lfc_cutoff) %>% as_tibble
}else{
  message("Cannot find significant windows in this dataset. Try lowering the p-value and log2FoldChange thresholds!")
}
```

```{r extractRegion,eval=FALSE}
if(sigWindows>0){
	resultRegions <- extractRegions(windowRes  = resultWindows, padjCol    = padjCol, padjThresh = p_value_cutoff, log2FoldChangeThresh = lfc_cutoff) %>% as_tibble
}
```

                     
### Writing Bed file

```{r writing bed file}
if(sigWindows>1){
	toBED(windowRes = resultWindows, regionRes = resultRegions,padjThresh=p_value_cutoff,
	  	padjCol   = padjCol, fileName  = output_bed_file)
}else{
  message("This analysis does not have enough <=2 significant windows to create BED file for visualization")
}
```


## Save Session
```{r save_data}
# save enriched windows, gzip results file if the file suffix is .gz
if(grepl("\\.gz$",output_windows_file,ignore.case = TRUE)){
  gz_out <- gzfile(output_windows_file,"w")
  write.table(resultWindows,file=gz_out,sep="\t",quote=FALSE,row.names=FALSE,col.names=TRUE)
  close(gz_out)
}else{
  write.table(resultWindows,file=output_windows_file,sep="\t",quote=FALSE,row.names=FALSE,col.names=TRUE)
}
# save enriched regions
if(sigWindows>0){
  if(grepl("\\.gz$",output_regions_file,ignore.case = TRUE)){
    gz_out <- gzfile(output_regions_file,"w")
    write.table(resultRegions,file=gz_out,sep="\t",quote=FALSE,row.names=FALSE,col.names=TRUE)
    close(gz_out)
  }else{
    write.table(resultRegions,file=output_regions_file,sep="\t",quote=FALSE,row.names=FALSE,col.names=TRUE)
  }
}
# save session
# Warning! session images can be heavy!
if(nchar(output_Rdata)>5){
  save.image( file = output_Rdata)
}
```

## Session Info

```{r session info}
sessionInfo()
```
